---
title: "Syntax and Output for Analyses"
author: "Ian Ruginski"
date: "3/14/2019"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false  
    theme: readable
---
This is output and analysis for reproducability for analyses included in the paper, "GPS-use negatively affects environmental learning through spatial transformation abilities," authored by Ian T. Ruginski, Sarah H. Creem-Regehr, Jeanine K. Stefanucci, and Elizabeth Cashdan. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)

```

### Data Cleaning & Loading Packages

Read in the data file. Load library packages. 

```{r, message = FALSE}
#loading libraryd packages 
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
options(digits=2)
library(psych)
library(car)
library(lavaan)
library(tidylog)
library(Hmisc)

#read in data
clean <- read.csv("ian_diss_cleaneddata.csv")
pt <- read.csv("PerspectiveTakingDataRaw.csv")

#cleaning PT data 
#NOTE: If a participant leaves an item blank, they should be assigned an angular deviation of 90 degrees for that item (90 represents chance performance, given that the angular deviation can vary from 0 to 180) (this was done when entering data)

#If the angular deviation on any item exceeds 180, it should be subtracted from 360 to determine the smallest deviation possible. This is because the angular deviation between two angles does not exceed 180 degrees. 

pt$SOTe1c <- ifelse(pt$SOTe1>=180, 360-pt$SOTe1, pt$SOTe1)
pt$SOTe2c <- ifelse(pt$SOTe2>=180, 360-pt$SOTe2, pt$SOTe2)
pt$SOTe3c <- ifelse(pt$SOTe3>=180, 360-pt$SOTe3, pt$SOTe3)
pt$SOTe4c <- ifelse(pt$SOTe4>=180, 360-pt$SOTe4, pt$SOTe4)
pt$SOTe5c <- ifelse(pt$SOTe5>=180, 360-pt$SOTe5, pt$SOTe5)
pt$SOTe6c <- ifelse(pt$SOTe6>=180, 360-pt$SOTe6, pt$SOTe6)
pt$SOTe7c <- ifelse(pt$SOTe7>=180, 360-pt$SOTe7, pt$SOTe7)
pt$SOTe8c <- ifelse(pt$SOTe8>=180, 360-pt$SOTe8, pt$SOTe8)
pt$SOTe9c <- ifelse(pt$SOTe9>=180, 360-pt$SOTe9, pt$SOTe9)
pt$SOTe10c <- ifelse(pt$SOTe10>=180, 360-pt$SOTe10, pt$SOTe10)
pt$SOTe11c <- ifelse(pt$SOTe11>=180, 360-pt$SOTe11, pt$SOTe11)
pt$SOTe12c <- ifelse(pt$SOTe12>=180, 360-pt$S2Te10, pt$SOTe12)

pt <- select(pt, SOTe1c:SOTe12c) #subset newly coded items
pt$SOTmean <- rowMeans(select(pt, SOTe1c:SOTe12c))

#add PT data to full dataframe
clean <- cbind(clean, pt$SOTmean) 
#change names
colnames(clean)[87] <- "SOTmean"

#make aggregate distance/pointing measures
clean$Pointing <- (clean$BetweenRtPt + clean$WithinRtPt) / 2
clean$Distance <- (clean$BetweenDist + clean$WithinDist) / 2

#make GPS variable numeric
clean$GPSuse <- as.numeric(clean$Q14)

#below code writes clean datafile
#write.csv(clean, "ian_diss_cleaneddatawithpt.csv")

#add bidimensional regression cognitive map data
bidim <- read.csv("bidimregression_results.csv")
bidimrsq <- select(bidim, id, r.2) #extract rsquared column only (variance explained in cog maps)
cleanr <- merge(clean, bidimrsq, by="id")#make new dataframe w bidim data

#make inverse scores so that higher means better ability for all variables consistently
cleanr$SOTmeanI <- -cleanr$SOTmean 
cleanr$BetweenRtPtI <- -cleanr$BetweenRtPt
cleanr$WithinRtPtI <- -cleanr$WithinRtPt
cleanr$DistanceI <- -cleanr$Distance
```

### Visualizing Distribution of GPS-use, Preliminary Analyses, & Assumptions
In this section we visualize distributions of variables and assess multicollinearity for mental rotation and perspective-taking, which are known to be very highly correlated. We also determine if GPS-use affects spatial transformation and environmental learning variables, to determine if mediation is a possibility apriori before testing our primary hypotheses. 


#### Descriptives & correlation matrix
```{r message=FALSE}
library(moments)

#make APA style Descriptives table
statdesc <- cleanr %>%
  summarise_at(vars(SBtotal, GPSuse, MRT, SOTmean, WithinRtPt:BetweenRtPt, 
                    Distance:r.2, Q1, Q4, UTdist),
                    list(~mean(.), ~sd(.), ~skewness(.), ~kurtosis(.)), na.rm=TRUE) %>% 
    gather("Variable", "Value") %>%  
    separate(Variable, c("Variable", "Statistic"), "_") %>%
    spread("Statistic", "Value") %>% 
  select(Variable, mean, sd, skewness, kurtosis) %>% 
  arrange(factor(Variable, levels = c("SBtotal", "GPSuse", "MRT", "SOTmean", 
                                      "WithinRtPt", "BetweenRtPt", 
                                      "Distance", "r.2", "Q1", "Q4", "UTdist")))

kable(statdesc) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
write.table(format(statdesc, digits=2), file = "statdesc_diss.txt", sep = ",", quote = FALSE, row.names = F)

#correlations
cor.df <- cleanr %>%
  select(SBtotal, GPSuse, MRT, SOTmean, WithinRtPt:BetweenRtPt, Distance:r.2, Q1, Q4, UTdist) %>% 
  as.matrix() %>%
  rcorr(type = "spearman")
options(scipen=10)

rcors <- kable(cor.df$r) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
write.table(formatC(cor.df$r, digits=2), 
            file = "rcors_diss.txt", sep = ",", quote = FALSE, row.names = T)
rcors

pcors <- kable(cor.df$P) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
pcors
write.table(formatC(cor.df$P, digits=2, format="g"),
            file = "pcors_diss.txt", sep = ",", quote = FALSE, row.names = T)

 
```
Correlations (r) displayed before p-values for each correlation. 


#### Sex differences
```{r}
cleanr$GenderC <- factor(cleanr$Gender, levels=c("Female", "Male"))

sex.t <- cleanr %>% 
  select(SBtotal, GPSuse, MRT, 
         SOTmean, WithinRtPt:BetweenRtPt, 
         Distance:r.2, Q1, Q4, UTdist) %>%
  map(~tidy(t.test(. ~ cleanr$GenderC))) %>% 
  bind_rows() %>% 
  mutate(Variable = c("SBtotal", "GPSuse", "MRT", "SOTmean",
                "WithinRtPt", "BetweenRtPt",
                "Distance", "r.2", "Q1", "Q4", "UTdist"))
kable(sex.t) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))


sex.d <- cleanr %>% 
  select(GenderC, SBtotal, GPSuse, MRT, 
         SOTmean, WithinRtPt:BetweenRtPt, 
         Distance:r.2, Q1, Q4, UTdist) %>%
  cohen.d(., "GenderC") %>% .$cohen.d

kable(sex.d) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))


statdesc.sex <- cleanr %>% filter(GenderC != "NA") %>% 
  group_by(GenderC) %>% 
  summarise_at(vars(SBtotal, GPSuse, MRT, SOTmean, WithinRtPt:BetweenRtPt, 
                    Distance:r.2, Q1, Q4, UTdist),
                    list(~mean(.), ~sd(.)), na.rm=TRUE)

kable(statdesc.sex) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

#### Visualization of GPS use distribution
```{r, echo=TRUE}
#lets look at the variable distributions determine estimator... probably have to use MLR due to GPS-use
#get descriptives
describe(cleanr$GPSuse)
#plot GPS counts as well
theme_set(theme_classic(base_size=18))
ggplot(cleanr, aes(x=as.factor(GPSuse), fill=as.factor(GPSuse))) + geom_bar(alpha=.8) +
  labs(title="Count Plot", 
       subtitle="GPS Use",
       x="Self-reported GPS Use") + guides(fill=guide_legend(title="Likert scale")) + scale_y_continuous(breaks=seq(0, 80, 15)) + ylab("Number of participants") + scale_x_discrete(labels=c("1" = "Never", "2" = "Rarely", "3" = "Sometimes", "4" = "Often", "5" = "Always"))
#ggsave("GPSuse_distribution.png", width=8, height=6)
```

#### Multicollinearity
Assessing multicollinearity between transformation abilities..
```{r, echo=TRUE}
btwmodel <- lm(WithinRtPt ~ MRT + SOTmeanI, cleanr)
vif(btwmodel)
```


#### Preliminary GPS models - predicting outcomes
Preliminary models of GPS-use on outcomes to determine if mediation makes sense apriori.
```{r, echo=TRUE, warning=FALSE}
#GPS preliminary models
model.gps.env <- 'EnvLearn =~ NA*BetweenRtPtI + WithinRtPtI + DistanceI + r.2
EnvLearn ~~ 1*EnvLearn
EnvLearn~GPSuse + Q1 + Q4 + UTdist'
fit.gps.env <- sem(model.gps.env, data = cleanr, missing="FIML", estimator="MLR")
summary(fit.gps.env, standardized=TRUE)

model.gps.mrt <- '
MRT~GPSuse + Q1 + Q4 + UTdist'
fit.gps.mrt <- sem(model.gps.mrt, data = cleanr,missing="FIML", estimator='MLR')
summary(fit.gps.mrt, standardized=TRUE)

model.gps.pt <- '
SOTmean~GPSuse + Q1 + Q4 + UTdist'
fit.gps.pt <- sem(model.gps.pt, data = cleanr,missing="FIML", estimator='MLR')
summary(fit.gps.pt, standardized=TRUE)

model.gps.mob <- '
GPSuse~Q1 + Q4+UTdist'
fit.gps.mob <- sem(model.gps.mob, data = cleanr,missing="FIML", estimator='MLR')
summary(fit.gps.mob, standardized=TRUE)
```

### Measurement Model
Assesing fit for the measurement model of the visual environmental learning latent factor. 
```{r, echo=TRUE, warning=FALSE}
model.measurement <- 'EnvLearn =~ NA*BetweenRtPt + WithinRtPt + r.2 + Distance
EnvLearn ~~ 1*EnvLearn'
fit.measurement <- sem(model.measurement, data = cleanr, missing="FIML", estimator="ML")
summary(fit.measurement, fit.measures = TRUE, standardized=TRUE)

#fits pretty well & all observed vars load
```
### Structural Equation Models and Model Comparison

#### Model 1: Mental rotation does not predict perspective taking
```{r model1, echo=TRUE, warning=FALSE}
#model 1: less saturated (no path from MRT to PT)
model.lesssat <- 'EnvLearn =~ NA*BetweenRtPtI + WithinRtPtI + DistanceI + r.2
SOTmeanI ~ GPSuse + 0*MRT + SBtotal
MRT ~ GPSuse + SBtotal
GPSuse ~ SBtotal
EnvLearn ~ MRT + SOTmeanI + GPSuse + SBtotal
EnvLearn ~~ 1*EnvLearn'
fit.lesssat <- sem(model.lesssat, data = cleanr, estimator="MLR", missing="FIML")
summary(fit.lesssat, fit.measures=TRUE, standardized=TRUE)
```

#### Model 2: Perspective-taking mediates the effect of mental rotation on environmental learning
```{r model2, echo=TRUE, warning=FALSE}
#model 2: more saturated (path from MRT to PT)
model.moresat <- 'EnvLearn =~ NA*BetweenRtPtI + WithinRtPtI + DistanceI + r.2
SOTmeanI ~ GPSuse + MRT + SBtotal
MRT ~ a*GPSuse + SBtotal
GPSuse ~ SBtotal
EnvLearn ~ MRT + SOTmeanI + GPSuse + SBtotal
EnvLearn ~~ 1*EnvLearn'
fit.moresat <- sem(model.moresat, data = cleanr, estimator="MLR", missing="FIML")
summary(fit.moresat, fit.measures=TRUE, standardized=TRUE)

#model 2b: more saturated, but diff direction (path from transforms to GPS)
#shows that directionality does not matter
model.moresat.altdir <- 'EnvLearn =~ NA*BetweenRtPtI + WithinRtPtI + DistanceI + r.2
SOTmeanI ~ MRT + SBtotal
MRT ~ SBtotal
GPSuse ~ SBtotal + MRT + SOTmeanI
EnvLearn ~ MRT + SOTmeanI + GPSuse + SBtotal
EnvLearn ~~ 1*EnvLearn'
fit.moresat.altdir <- sem(model.moresat.altdir, data = cleanr, estimator="MLR", missing="FIML")
summary(fit.moresat.altdir, fit.measures=TRUE, standardized=TRUE)
```

#### Comparison of Models 1 & 2
Though model 2 appears to fit the data better based on some fit indices, we will used an adjusted Satorra-Bentler Chi-squared difference test to explicitly compare the two nested models.
```{r model.comparison, echo=TRUE}
lavTestLRT(fit.lesssat, fit.moresat) #if p < .05, model different from data. so we want to keep more saturated model with addtl parameter estimated which in this case is path from MRT to PT
```
We find that the more saturated model (where mental rotation predicts perspective-taking, model 2) more accurately represents the data.

###Testing Mediation by Bootstrapping Indirect Effects
Now that we have selected the best fitting model, we will bootstrap indirect effects. Evidence of mediation is present if the bias-corrected CI does not contain zero for an indirect effect. We will also bootstrap difference tests between some direct and indirect effects to evaluate which variables contribute relatively more to spatial transformation and environmental learning ability outcomes. 
```{r model.med, echo=TRUE, warning=FALSE}
set.seed(28) #for reproducability of bootstrapping analyses
model.boot <- 'EnvLearn =~ NA*BetweenRtPtI + WithinRtPtI + DistanceI + r.2
SOTmeanI ~ c*GPSuse + e*MRT + h*SBtotal
MRT ~ a*GPSuse + i*SBtotal
GPSuse ~ f*SBtotal
EnvLearn ~ b*MRT + d*SOTmeanI + g*GPSuse + j*SBtotal
EnvLearn ~~ 1*EnvLearn
#calculate indirect effect. a* is where our predictor predicts the mediating variable.
#b* is where our mediator predicts the final dependent variable.
ab := a*b #Indirect effect of GPS on Spatial Learning through MRT
cd := c*d #Ind of GPS on SL through Perspective Taking
de := d*e #Ind of MRT on SL through Perspective Taking
ae := a*e #Ind of GPS on PT through MRT
aed := a*e*d #Ind of GPS on SL through MRT and PT
diff1 := h - ae #Diff between SBSOD direct on PT and GPS to PT through MRT
diff2 := h - c #Diff between effects of SB and GPS on PT
diff3 := a - i #Diff between effects of SB and GPS on MRT
diff4 := j - aed #Difference between SBSOD direct on SL and Indirect from GPS to SL'
fit.boot <- sem(model.boot, data = cleanr, se="bootstrap", test="bootstrap", bootstrap=2000)
```

Check if bias-corrected confidence intervals of bootstrapped indirect effects contain zero for evidence of mediation. This will also provide values for the difference in effects tests.
```{r model2_indeffects, echo=TRUE, warning=FALSE}
#bootstrapping  bias corrected CI's for indirect effects
parest <- parameterEstimates(fit.boot, boot.ci.type = "bca.simple", standardized = TRUE)
#pull out parameters of interest
ind_table <- parest %>%
  select(label,est,std.all,se, ci.lower:ci.upper)
ind_table
```

